Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus

### 2024-01-31 manuscript_thev_transcriptome.Rmd
added styling to word output

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document: default
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(tibble)
library(scales)
library(flextable)
library(english)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(tab.cap.pre = "", tab.cap.sep = "")

@@

Hemorrhagic enteritis (HE) is a disease affecting 6-12-week-old turkeys characterized by immunosuppression (IS) and bloody diarrhea. This disease is caused by *Turkey Hemorrhagic Enteritis Virus* (THEV) of which avirulent strains (THEV-A) that do not cause HE but retain the immunosuppressive ability have been isolated. The THEV-A Virginia Avirulent Strain (VAS) is still used as a live vaccine despite its immunosuppressive properties. We have performed the first RNA-sequencing experiment characterizing THEV’s transcriptome, yielding the most detailed insight into THEV gene expression, to set the stage for further experimentation with specific viral genes that may mediate IS. After infecting a turkey B-cell line (MDTC-RP19) with the VAS vaccine strain, samples in triplicates were collected at 4-, 12-, 24-, and 72-hours post-infection. Total RNA was subsequently extracted, and poly-A-tailed mRNA sequencing done. After trimming the raw sequencing reads with the `Trim-galore`, reads were mapped to the THEV genome using `Hisat2` and transcripts assembled with `StringTie`. We identified 29 transcripts from our RNA-seq data all of which consisted of novel exons albeit some exons matched the predicted ORFs. The three predicted splice junctions were also corroborated in our data. We performed PCR amplification of THEV cDNA, cloned the PCR products, and Sanger sequencing was used to validate all identified splice junctions. During validation, we identified 5 additional transcripts some of which were further validated by 3'RACE data. Thus, the transcriptome of THEV consists of 34 unique transcripts with the coding capacity for all predicted ORFs. However, we found 8 predicted ORFs to be incomplete as either an upstream, in-frame start codon was identified or additional coding exons were found, making the actual expressed versions of these ORFs longer. We also identified 7 novel unpredicted ORFs that could be encoded by some transcripts; albeit it is beyond the scope of this manuscript to investigate whether they are indeed expressed. In keeping with all Adenoviruses, our data shows that all THEV transcripts are spliced, and organized in transcription units under the control of their cognate promoter.
\newpage

@@

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, and direct genome packaging. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 genes, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Aboezz *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection(h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

@@

Considering individual mRNAs, TRXPT_21 -- from the E2 region -- was the most significantly expressed at 12 h.p.i, constituting about `r abn_each_12`% of the total expression of all transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region transcripts -- TRXPT_10 and TRXPT_14 were the most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions at each time point using the raw reads. For individual junctions, we counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLTU was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junction reads (**Table 2a** and **Supplementary Table 1a**). The top most abundant junctions at 12 h.p.i remained the most significantly expressed at 24 h.p.i also. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b** and **Supplementary Table 1b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP region junctions were not only the most abundant overall -- accounting for `r round(most_sig_72, 1)`% of all junction reads, -- but also contained the most significantly expressed individual junctions (**Table 2c**, **Supplementary Table 1c** and **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).

@@

**Early Region 1 (E1) transcripts**    
This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. In MAdVs, this region is subdivided into E1a and E2b units but the transcripts found in our data categorized under this region do not appear to so divided.

@@

While both TRXPT_6 and TRXPT_7 have the CP for Ad-pol with secSSC usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are encoded by separate mRNAs with identical first three 5' exons and TTS, but the splice junction to the terminal exons are different. We checked for a longer splice junction between the third and fourth (terminal) exons of TRXPT_7 with our junction validation method (targeted PCR, cloning, and Sanger sequencing) and discovered a unique splice junction (10,981-7062bp) not found in our RNA-seq data. If initiated from the E2 TSS and terminated at the E2 TTS, this transcript(TRXPT_31) would encode Ad-pol exactly as predicted as its 5'-most CDS (**Figure 7**).

@@

We identified seven novel transcripts here (`r paste0("TRXPT_", c(22:27, 29), collapse = ", ")`) from our RNA-seq data, all originating from two distinct TSSs -- we consider the first TSS (position 18,230bp) as corresponding to the L4P and the other at 18,727bp as corresponding to the E3 promoter (E3P). These E3 transcripts collectively have the CP for several predicted THEV ORFs: 100K, 22K, 33K, pVIII, and E3, as well as Fiber (IV) and ORF7 belonging to the MLTU. But some of these CDSs are different than predicted due to either unknown exons or the presence of an in-frame upstream SSC. For instance, 33K is one of the few THEV ORFs predicted to be spliced from two exons; however, we discovered a significantly longer four-exon ORF (e33K) on TRXPT_24 that contains it almost entirely. The first two exons of e33K were not predicted but the last two match the predicted exons and the CDS is in-frame, albeit the first 20bp of the predicted 33K (including the SSC at 20,142bp) is spliced out as part of the second intron of TRXPT_24. Thus, the bona fide 33K (e33K) is a 19.8 kDa, 171 residue protein spanning four exons instead of the predicted 120 aa protein. TRXPT_24 also has the CP for pVIII and E3 if we consider downstream SSC usage. However, the predicted E3 has an upstream in-frame SSC; thus this longer version of E3 (eE3) is the genuinely expressed ORF. TRXPT_29 is the shortest transcript in this TU. It is a two-exon transcript, both exons comprising the CDS. The product of TRXPT_29 is a novel 73 residue protein (8.3KI) sharing the SSC of e33K but with a unique STC. TRXPT_23 being spliced identically as TRXPT_29 also encodes 8.3KI from its first SSC. Similarly, TRXPT_22 also encodes a 73 aa novel protein (8.3KII) from its first SSC that shares over 80% similarity with 8.3KI, but it differs from 8.3KI at the C-terminus. Considering downstream SSC usage, both TRXPT_22 and TRXPT_23 can encode pVIII and eE3 in that order, but TRXPT_23 being longer, has the CP for the Fiber ORF also.

As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 essentially share the same genomic space, their validation was done with a single primer pair and they were differentiated from each other by cloning and Sanger sequencing (**Supplementary PCR methods**). In addition to corroborating the splice junctions for the aforementioned transcripts, the Sanger sequencing results also showed another splice  variant undetected in our RNA-seq transcriptome. This was a three-exon transcript (TRXPT_30) with its first and last exons spliced identically as TRXPT_23, but which also has the second exon of TRXPT_24 (**Figure 8**). The first CDS on TRXPT_30 spans all three exons, producing a novel 140 residue, 15.7kDa protein. Interestingly, the last 81 C-terminus residues of this new protein (e22K) are identical to 22K (89 residues), which is a single-exon ORF predicted to use the same SSC as 33K (20,142bp). Just as seen for 33K, all the transcripts in this region exclude the first 20bp of 22K (including the SSC) as part of their introns; therefore, the first 7 residues of 22K are lacking in e22K due to splicing. Hence, we consider e22K as a long variant of the predicted 22K ORF. Albeit the TSS and TTS of TRXPT_30 was not seen, we presume that they are similar to TRXPT_23, in which case it would also have the downstream CP of TRXPT_23.

@@

**Early Region 4 (E4) transcripts**   
This TU is found at the tail-end (3'-end) of the genome  and expressed from the anti-sense strand. Based on nucleotide position, ORF7 and ORF8 were predicted in this region [@Davison2003]; however, as ORF7 is neither on the same strand as ORF8 nor transcribed from a promoter in the E4 region, only ORF8 can legitimately be classified as a transcript in this TU. This is corroborated by our RNA-seq data, as only one transcript was identified in this region on the anti-sense strand (**Figure 9**). The transcript (TRXPT_28) spans 25192-26247bp and is spliced at 25701-26055bp, making a two-exon transcript. The second exon fully matches the predicted ORF8 with 12 extra base pairs at the 3'-end. However, there is a SSC in the first exon at position 26246bp (192bp upstream of the predicted SSC). The encoded protein from this SSC is in-frame with the predicted SSC found in the second exon; hence, we consider this protein (eORF8 -- 26.4 kDa, 229 aa), a longer isoform of the predicted ORF8, the genuinely expressed ORF with an identical C-terminus to the predicted ORF8 protein.

@@

The use of short read deep sequencing to reconstruct full AdV mRNA structures provides excellent results, especially for mapping the splice sites. However, due to the substantial overlapping nature of AdV mRNAs coupled with the fragmentation step in the library preparation protocol, mapping the precise TSS and TTS of the assembled transcripts is difficult. Also, similar transcripts with substantial overlaps may be assembled as one longer mRNA, since the short reads alone do not provide enough context for the transcript assembler (`StringTie`) to distinguish them. In our results, we see transcripts in the same TU initiated or terminated in the same approximate area (10-70bp and 1-300bp apart for TSS and TTS, respectively) but not precisely at the same position. We consider the most upstream TSS or most downstream TTS for the transcripts involved but we present them unchanged in all the figures shown. Also, by comparison to the more well-studies MAdV transcriptomes, we think that a few long transcripts in the MLTU (TRXPT_9, TRXPT_10, and TRXPT_11) are probably a result of fusing some L4P-derived transcripts to the terminal exons of the bona fide MLTU transcripts by StringTie, making them significantly longer. These mRNAs do not only have unusually many exons for an AdV, but their last three or four exons are also identical to the L4P-derived mRNAs. Future studies using long read sequencing technologies are necessary to provide conclusive data for precisely mapping the TSS and TTS, as well as teasing apart the bona fide structures of the long MLTU transcripts. Furthermore, it is not unreasonable to presume that several splice variants were undiscovered in our work as evidenced firstly by finding unique transcripts using 3'RACE and during our splice junction validation steps. And secondly, recent studies [@Donovan2020; @Zhao2014; @Westergren2021] are still discovering novel mRNA variants for even the best studied MAdVs decades later. Another observation made is that all the TTSs in THEV's transcriptome are in close proximity to A/T-rich sequences which we presume to be polyadenylation signal sequences (PASS). Interestingly, some of these PASSs are located in the immediate vicinity of two closely-located TTSs expressed on opposite strands. Namely, the E1 and E2B/IM TTSs have an almost palindromic PASS between them, as do the E4 (anti-sense strand) and the sense strand TRXPT_12 and TRXPT_27.

@@

Data is available on request by contacting the designated corresponding author

### CODE AVAILABILITY
All the code/scripts in the entire analysis pipeline are available on github (<https://github.com/Abraham-Quaye/thev_transcriptome>) 
\newpage

@@

`LC Sciences` - RNA sequencing was done here   
`Eton Bioscience, Inc, San Diego, CA` - All Sanger sequencing validations was done here 
`BYU high performance computing systems` - Memory intensive analysis were run here.
\newpage

>>>>

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  word_document:
    reference_docx: style_word_output.docx
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(tibble)
library(scales)
library(flextable)
library(english)
library(rticles)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(tab.cap.pre = "", tab.cap.sep = "")

@@

We have performed an RNA-sequencing experiment characterizing the transcriptome of turkey hemorrhagic enteritis virus (THEV) for the first time, yielding the only insight into THEV's gene expression patterns. Previously, THEV's genome had been predicted to encode 23 open reading frames (ORFs). In this work we identified 29 transcripts from our RNA-seq data all of which consisted of novel exons albeit some exons matched the predicted ORFs. The three predicted splice junctions were also corroborated by our data. We performed PCR amplification of THEV cDNA and cloned the PCR products, and Sanger sequencing was used to validate all identified splice junctions. During validation, we identified 5 additional transcripts some of which were further validated by 3'RACE data. Thus, the transcriptome of THEV consists of 34 unique transcripts with the coding capacity for all predicted ORFs. However, we found 6 predicted ORFs (ORF1, E3, 33K, ORF8, IVa2, and protease) to be truncated predictions as either an in-frame upstream start codon was identified or additional coding exons were found. We also identified 3 predicted ORFs with longer or shorter isoforms, and 7 novel unpredicted ORFs that could be encoded by some transcripts; albeit it is beyond the scope of this manuscript to investigate whether they get translated. Similar to other adenoviruses, THEV also produce multiple distinctly spliced transcripts that code for the same protein across its genome. Also, our data shows that all THEV transcripts are spliced, and organized in five transcription units under the control of their cognate promoter. However, our data suggests that THEV's temporal regulation may be different from other adenoviruses. Over 2,300 unique splice junctions were found across the genome, mostly at low levels. This low-level use of broad alternative splicing patterns is thought to enable the virus to maximize its coding potential in an evolving environment. 
\newpage

@@

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, and direct genome packaging. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 genes, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Aboezz *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection (h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example, all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

@@

Considering individual mRNAs, TRXPT_21 -- from the E2 region -- was the most significantly expressed at 12 h.p.i, constituting about `r abn_each_12`% of the total expression of all transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region transcripts -- TRXPT_10 and TRXPT_14 were the most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions at each time point using the raw reads. For individual junctions, we counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLTU was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junction reads (**Table 2a** and **Supplementary Table 1a**). The topmost abundant junctions at 12 h.p.i remained the most significantly expressed at 24 h.p.i also. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b** and **Supplementary Table 1b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP region junctions were not only the most abundant overall -- accounting for `r round(most_sig_72, 1)`% of all junction reads, -- but also contained the most significantly expressed individual junctions (**Table 2c**, **Supplementary Table 1c** and **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).

@@

**Early Region 1 (E1) transcripts**    
This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. In MAdVs, this region is subdivided into E1a and E2b units, but the transcripts found in our data categorized under this region do not appear to so divided.

@@

While both TRXPT_6 and TRXPT_7 have the CP for Ad-pol with secSSC usage, in all AdVs studied, the two proteins (pTP and Ad-pol) are encoded by separate mRNAs with identical first three 5' exons and TTS, but the splice junction to the terminal exons are different. We checked for a longer splice junction between the third and fourth (terminal) exons of TRXPT_7 with our junction validation method (targeted PCR, cloning, and Sanger sequencing) and discovered a unique splice junction (10,981-7062bp) not found in our RNA-seq data. If initiated from the E2 TSS and terminated at the E2 TTS, this transcript (TRXPT_31) would encode Ad-pol exactly as predicted as its 5'-most CDS (**Figure 7**).

@@

We identified seven novel transcripts here (`r paste0("TRXPT_", c(22:27, 29), collapse = ", ")`) from our RNA-seq data, all originating from two distinct TSSs -- we consider the first TSS (position 18,230bp) as corresponding to the L4P and the other at 18,727bp as corresponding to the E3 promoter (E3P). These E3 transcripts collectively have the CP for several predicted THEV ORFs: 100K, 22K, 33K, pVIII, and E3, as well as Fiber (IV) and ORF7 belonging to the MLTU. But some of these CDSs are different than predicted due to either unknown exons or the presence of an in-frame upstream SSC. For instance, 33K is one of the few THEV ORFs predicted to be spliced from two exons; however, we discovered a significantly longer four-exon ORF (e33K) on TRXPT_24 that contains it almost entirely. The first two exons of e33K were not predicted but the last two match the predicted exons and the CDS is in-frame, albeit the first 20bp of the predicted 33K (including the SSC at 20,142bp) is spliced out as part of the second intron of TRXPT_24. Thus, the bona fide 33K (e33K) is a 19.8 kDa, 171 residue protein spanning four exons instead of the predicted 120 aa protein. TRXPT_24 also has the CP for pVIII and E3 if we consider downstream SSC usage. However, the predicted E3 has an upstream in-frame SSC; thus, this longer version of E3 (eE3) is the genuinely expressed ORF. TRXPT_29 is the shortest transcript in this TU. It is a two-exon transcript, both exons comprising the CDS. The product of TRXPT_29 is a novel 73 residue protein (8.3KI) sharing the SSC of e33K but with a unique STC. TRXPT_23 being spliced identically as TRXPT_29 also encodes 8.3KI from its first SSC. Similarly, TRXPT_22 also encodes a 73 aa novel protein (8.3KII) from its first SSC that shares over 80% similarity with 8.3KI, but it differs from 8.3KI at the C-terminus. Considering downstream SSC usage, both TRXPT_22 and TRXPT_23 can encode pVIII and eE3 in that order, but TRXPT_23 being longer, has the CP for the Fiber ORF also.

As the splice junctions of TRXPT_22, TRXPT_23, TRXPT_24, and TRXPT_29 essentially share the same genomic space, their validation was done with a single primer pair, and they were differentiated from each other by cloning and Sanger sequencing (**Supplementary PCR methods**). In addition to corroborating the splice junctions for the aforementioned transcripts, the Sanger sequencing results also showed another splice  variant undetected in our RNA-seq transcriptome. This was a three-exon transcript (TRXPT_30) with its first and last exons spliced identically as TRXPT_23, but which also has the second exon of TRXPT_24 (**Figure 8**). The first CDS on TRXPT_30 spans all three exons, producing a novel 140 residue, 15.7kDa protein. Interestingly, the last 81 C-terminus residues of this new protein (e22K) are identical to 22K (89 residues), which is a single-exon ORF predicted to use the same SSC as 33K (20,142bp). Just as seen for 33K, all the transcripts in this region exclude the first 20bp of 22K (including the SSC) as part of their introns; therefore, the first 7 residues of 22K are lacking in e22K due to splicing. Hence, we consider e22K as a long variant of the predicted 22K ORF. Albeit the TSS and TTS of TRXPT_30 was not seen, we presume that they are similar to TRXPT_23, in which case it would also have the downstream CP of TRXPT_23.

@@

**Early Region 4 (E4) transcripts**   
This TU is found at the tail-end (3'-end) of the genome  and expressed from the anti-sense strand. Based on nucleotide position, ORF7 and ORF8 were predicted in this region [@Davison2003]; however, as ORF7 is neither on the same strand as ORF8 nor transcribed from a promoter in the E4 region, only ORF8 can legitimately be classified as a transcript in this TU. This is corroborated by our RNA-seq data, as only one transcript was identified in this region on the anti-sense strand (**Figure 9**). The transcript (TRXPT_28) spans 25192-26247bp and is spliced at 25701-26055bp, making a two-exon transcript. The second exon fully matches the predicted ORF8 with 12 extra base pairs at the 3'-end. However, there is an SSC in the first exon at position 26246bp (192bp upstream of the predicted SSC). The encoded protein from this SSC is in-frame with the predicted SSC found in the second exon; hence, we consider this protein (eORF8 -- 26.4 kDa, 229 aa), a longer isoform of the predicted ORF8, the genuinely expressed ORF with an identical C-terminus to the predicted ORF8 protein.

@@

The use of short read deep sequencing to reconstruct full AdV mRNA structures provides excellent results, especially for mapping the splice sites. However, due to the substantial overlapping nature of AdV mRNAs coupled with the fragmentation step in the library preparation protocol, mapping the precise TSS and TTS of the assembled transcripts is difficult. Also, similar transcripts with substantial overlaps may be assembled as one longer mRNA since the short reads alone do not provide enough context for the transcript assembler (`StringTie`) to distinguish them. In our results, we see transcripts in the same TU initiated or terminated in the same approximate area (10-70bp and 1-300bp apart for TSS and TTS, respectively) but not precisely at the same position. We consider the most upstream TSS or most downstream TTS for the transcripts involved but we present them unchanged in all the figures shown. Also, by comparison to the more well-studies MAdV transcriptomes, we think that a few long transcripts in the MLTU (TRXPT_9, TRXPT_10, and TRXPT_11) are probably a result of fusing some L4P-derived transcripts to the terminal exons of the bona fide MLTU transcripts by StringTie, making them significantly longer. These mRNAs do not only have unusually many exons for an AdV, but their last three or four exons are also identical to the L4P-derived mRNAs. Future studies using long read sequencing technologies are necessary to provide conclusive data for precisely mapping the TSS and TTS, as well as teasing apart the bona fide structures of the long MLTU transcripts. Furthermore, it is not unreasonable to presume that several splice variants were undiscovered in our work as evidenced firstly by finding unique transcripts using 3'RACE and during our splice junction validation steps. And secondly, recent studies [@Donovan2020; @Zhao2014; @Westergren2021] are still discovering novel mRNA variants for even the best studied MAdVs decades later. Another observation made is that all the TTSs in THEV's transcriptome are in close proximity to A/T-rich sequences which we presume to be polyadenylation signal sequences (PASS). Interestingly, some of these PASSs are located in the immediate vicinity of two closely located TTSs expressed on opposite strands. Namely, the E1 and E2B/IM TTSs have an almost palindromic PASS between them, as do the E4 (anti-sense strand) and the sense strand TRXPT_12 and TRXPT_27.

@@

Data is also available on request by contacting the designated corresponding author

## CODE AVAILABILITY
All the code/scripts in the entire analysis pipeline are available on github (<https://github.com/Abraham-Quaye/thev_transcriptome>) 
\newpage

@@

`LC Sciences` - RNA sequencing was done here   
`Eton Bioscience, Inc, San Diego, CA` - All Sanger sequencing validations was done here 
`BYU high performance computing systems` - Memory-intensive analysis were run here.
\newpage
